*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-15 Code/Manuscript_O2.Rmd
working on introduction

---
title: "Functional Analyses of Distinct Oxygen Concentrations and Light Niches of *Prochlorococcus marinus*"
author:
  - name: '**Mireille Savoie**'
    affil: 1
    main: true
    email: msavoie@mta.ca
  - name: Julie Nadeau
    affil: 1
  - name: Laurel Genge
    affil: 1
  - name: Sylwia Śliwińska-Wilczewska
    affil: 1,2
  - name: Maximilian Berthold
    affil: 1
  - name: Naaman Omar
    affil: 1
  - name: Douglas A. Campbell
    affil: 1
    main: false
    orcid: ''
    email: 'dcampbel@mta.ca'
affiliation:
  - num: 1
    address: Department of Biology, Mount Allison University
institute:
  - mta: 'Department of Biology, Mount Allison University, 53 York St., Sackville NB, Canada, E4L 1C9.'
date: "`r format(Sys.Date())`"
output:
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  bookdown::word_document2:
    code_folding: show
    reference_docx: KnitTemplate.docx
    keep_md: yes
    toc: TRUE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
always_allow_html: true
bibliography: [Manuscript_O2.bib, RPackages.bib]
csl: plos-one.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

@@

%%%% Insert GAM figures to show lab results confirming Ocean Protein Portal environment results %%%

@@

%%% TO DO Fits and ANOVA %%
```{r MED4PAR, fig.cap = MED4PAR_cap, fig.height = 6, fig.width = 8, echo = FALSE}

%%% TO DO Fits and ANOVA %%
```{r SS120PAR, fig.cap = SS120PAR_cap, fig.height = 6, fig.width = 8, echo = FALSE}

%%% TO DO Fits and ANOVA %%
```{r MIT9313PAR, fig.cap = MIT9313PAR_cap, fig.height = 6, fig.width = 8, echo = FALSE}

@@

>>>>

---
title: "*Prochlorococcus marinus* responses to light and oxygen" 
author:
  - name: '**Mireille Savoie**'
    affil: 1
    main: true
    email: msavoie@mta.ca
  - name: Aurora Mattison
    affil: 1,2
  - name: Julie Nadeau
    affil: 1
  - name: Laurel Genge
    affil: 1
  - name: Sylwia Śliwińska-Wilczewska
    affil: 1
  - name: Maximilian Berthold
    affil: 1,3
  - name: Naaman Omar
    affil: 1
- name: Ondřej Prášil
    affil: 1,4
    main: false
  - name: Amanda M. Cockshutt
    affil: 1,5
    main: false
  - name: Douglas A. Campbell
    affil: 1
    main: false
    orcid: '0000-0001-8996-5463'
    email: 'dcampbel@mta.ca'
affiliation:
  - num: 1
    address: Department of Biology, Mount Allison University
institute:
  - mta: 'Department of Biology, Mount Allison University, 53 York St., Sackville NB, Canada, E4L 1C9.'
  - num: 2
institute:
  - ubc: 'Department of Biology, University of British Columbia'
  - num: 4
institute:
  - alga: 'MAX INSTITUTE'
  - num: 4
institute:
  - alga: 'Algatech, Czech Academy of Science'
  - num: 5
institute:
  - sfx: 'Department of Chemistry, St. Frances Xavier University'
date: "`r format(Sys.Date())`"
output:
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  bookdown::word_document2:
    code_folding: show
    reference_docx: KnitTemplate.docx
    keep_md: yes
    toc: TRUE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
always_allow_html: true
bibliography: [Manuscript_O2.bib, RPackages.bib]
csl: plos-one.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

## Aurora
*Prochlorococcus marinus*, a species of picocyanobacteria, are tiny, photosynthetic, marine prokaryotes. At nder 1 µm in diameter, they are the smallest photosynthetic organisms on the planet, but at 10^4^-10^5^ cells per millilitre of seawater  (ampbell and Vaulot, 1993), they heavily influence global biogeochemical cycles (Partensky et al., 1999) (Flombaum et al., 2013). *Prochlorococcus* are currently restricted to between 40°N and 40°S, from surface to 300 m  depth, spanning 3 orders of magnitudes of light (Partensky et al., 1999) (Chisholm et al., 1992). Some *Prochlorococcus* extend into Oxygen Minimum Zones (OMZ), at less than  20 µM O2 (Johnson et al., 1999). *Prochlorococcus* is divided into numerous strains, categorized into high-light clades HLI, HLII, HLIII or low-light clades LLI, LLII, LLIII, LLIV, LLV & LLVI, based on their 16S-23S intergenic transcribed ribosomal sequence signatures (Rocap et al., 2003). The LLIV clade, including strain MIT9313, falls near the base of the *Prochlorococcus* radiation. First isolated in 1993 from 135 m into the north Atlantic, Clade LLIV are characterized by preference for low light,  and are most populous between 120 m and 200 m depths (Partensky et al., 1999). LLIV members are also the only strains found in OMZ that have been, to date, cultured, although two as yet uncultured clades, LLV and LLVI, have been detected in the OMZ of the eastern Tropical Pacific Ocean (Lavin et al., 2010) ADDITIONAL REFS XXXX. LLII and LLIII clades are often grouped together as the second oldest phylogenetic clades and share an affinity for low light conditions; LLIII strain SS120, the first *Prochlorococcus* strain to be characterized in 1988, was isolated 135 m into the Sargasso Sea. The LLI clade includes NATL1A and NATL2A, which prefer moderate irradiances found between 30 and 100 m. HL clades are the most recently evolved and despite reduced genome size in comparison to LL ecotypes, have adapted to life at higher irradiances. HLII is the most abundant group in the North Atlantic and North Pacific Oceans, often constituting over 90% of the total population (Partensky et al., 1999),  and are most numerous around 50 m depth (Partensky et al., 1999). The HLI clade, including MED4, was found at the very surface of the Mediterranean Sea, just 5 m below the surface (Partensky et al., 1999).

*Prochlorococcus* are difficult to culture, as their reduced genome – the smallest of any known oxyphototroph – render them partially dependent upon mutualistic heterotrophic bacteria to  detoxify reactive oxygen species (Moore et al., 2007; Morris et al., 2008). MED4, SS120 and MIT9313 have been successfully cultured in laboratories, and used to show that ecotypic classifications correspond to biochemical differences among strains. CITATONS OUR LAB and OTHERS

The main target of photoinhibition is PSII, in particular its PsbA protein, which along with its homolog PsbD forms a heterodimer binding all redox active components at the photosystem’s reaction center core and is therefore the most vulnerable to redox damage (Barber, 2012). Under light-induced damage, PsbA transcripts accumulate as the organism attempts to replace the damaged protein, however protein levels remain relatively constant, if inoperative. The new copy cannot be incorporated until the damaged one is degraded thus, the degradation of PsbA has been considered the rate-limiting step in recovery from photoinhibition (Kanervo et al., 1993). The proteins believed to be responsible for this crucial step are the FtsH proteases, which have been found localized to the thylakoid membrane of cyanobacteria and higher plants (Sacharz et al., 2015) as well as co-purified with His-tagged PSII (Zak et al., 2001). Furthermore, mutants lacking FtsH have increased sensitivity to photoinactivation (Mann et al., 2000; Adam et al., 2005). 
Originally identified in Escherichia coli, (Tomoyasu et al., 1993) FtsH is a membrane-bound, ATP, zinc metalloprotease that extracts and degrades membrane-bound proteins for degradation (Chiba et al., 2002). FtsH are embedded in membranes via two N-terminal transmembrane domains connected by a small hydrophilic loop in the thylakoid lumen (Figure 9A). The main hydrophilic portion sits in the bacterial cytoplasm and contains an ATP-ase Associated with various cellular Activities (AAA) domain as well as a zinc-binding active site (Yoshioka-Nishimura and Yamamoto, 2014) (Figure 10a). Chloroplast FtsH were first identified when immunoblot analysis with antibodies against E. coli FtsH cross reacted with thylakoid membrane proteins in spinach (Lindahl et al., 1996). The chloroplast genome of higher plants can encode up to 8 FtsH homologs (Adam et al., 2005). Cyanobacterial genomes encode only 4 homologs. In the model cyanobacterium Synechocystis sp. PCC 6803 (Synechocystis) these are encoded by the slr1390, slr0228, slr1604 and sll1463 genes, which will be used to identify their respective proteins for clarity.  Slr0228 and sll1463 are found localized to patches in the thylakoid membrane (Sacharz et al., 2015) and it has been shown that sll1463 accumulation is dependent upon slr0228 (Pisareva et al., 2007). Furthermore, slr0228 has been shown to be indispensable for PSII repair and PSI biogenesis but intriguingly is not required for cell viability, nor is sll1463 (Mann et al., 2000). In contrast, slr1390 and slr1604 are found in both plasma and thylakoid membranes (Pisareva et al., 2007; Sacharz et al., 2015) and are essential for cell viability (Mann et al., 2000).  FtsH proteases have been shown to associate in homo- and hetero-oligomers (Yu et al., 2004). In Synechocystis, slr1390 and slr1604 associate in hetero-oligomers (Sacharz et al., 2015). Slr0228 and slr1604 form a hetero-hexameric ring in the thylakoid membrane with a central pore through which proteins are translocated as they are digested and it is this complex that is responsible for PSII repair (Adam et al., 2005; Boehm et al., 2012) (Figure 10b). Nixon et al. (2010) have proposed a model for the PSII repair process in which destabilization of PsbA, resulting in either unusual exposure or post-translational modification of PsbA’s N-terminus (Komenda et al., 2007), exposes an N-terminal “docking site” for FtsH to initiate  degradation (Figure 11).

Representatives of LL clades of *Prochlorococcus* suffer photoinhibition at irradiance levels preferred by their HL relatives. The main target of photoinhibition is PSII, in particular the PsbA protein, which along with its homolog PsbD forms a heterodimer binding all redox active components at the photosystem’s reaction center core and is therefore the most vulnerable to redox damage (Barber, 2012). Under light-induced damage, new PsbA cannot be incorporated until the damaged PsbA is removed thus, degradation of PsbA is a rate-limiting step in recovery from photoinhibition (Kanervo et al., 1993). PsbA degradation is mediated largely by FtsH proteases (Mann et al., 2000; Adam et al., 2005), localized to the thylakoid membrane (Sacharz et al., 2015) and PSII (Zak et al., 2001). FtsH is a membrane-bound, ATP, zinc metalloprotease that extracts and degrades membrane-bound proteins for degradation (Chiba et al., 2002) (Yoshioka-Nishimura and Yamamoto, 2014). In the model cyanobacterium *Synechocystis* sp. PCC 6803 4 isoforms of FtsH proteins are encoded by the slr1390, slr0228, slr1604 and sll1463 genes.  Slr0228 and sll1463 are found localized to patches in the thylakoid membrane (Sacharz et al., 2015) and it has been shown that sll1463 accumulation is dependent upon slr0228 (Pisareva et al., 2007). Furthermore, slr0228 has been shown to be indispensable for PSII repair and PSI biogenesis but intriguingly is not required for cell viability, nor is sll1463 (Mann et al., 2000). In contrast, slr1390 and slr1604 are found in both plasma and thylakoid membranes (Pisareva et al., 2007; Sacharz et al., 2015) and are essential for cell viability (Mann et al., 2000).  FtsH proteases have been shown to associate in homo- and hetero-oligomers (Yu et al., 2004). In Synechocystis, slr1390 and slr1604 associate in hetero-oligomers (Sacharz et al., 2015). Slr0228 and slr1604 form a hetero-hexameric ring in the thylakoid membrane with a central pore through which proteins are translocated as they are digested and it is this complex that is responsible for PSII repair (Adam et al., 2005; Boehm et al., 2012) (Figure 10b). Nixon et al. (2010) have proposed a model for the PSII repair process in which destabilization of PsbA, resulting in either unusual exposure or post-translational modification of PsbA’s N-terminus (Komenda et al., 2007), exposes an N-terminal “docking site” for FtsH to initiate  degradation (Figure 11).

*Prochlorococcus* genomes encode 4 FtsH proteins, henceforth referred to as FtsH1-4. Bonisteel et al. (2018) found that these were homologs of *Synechocystis* FtsH (Table 1).  Upon a shift to higher light MED4 upregulated expression of FtsH1 and FtsH2, homologous to *Synechocystis* slr0228 and slr1604, implicated in PSII repair. In contrast representative LL strain MIT9313 shows less expression of the FtsH proteases, and thus has fewer FtsH serving each photosystem,  MIT9313 expressed primarily FtsH3 and expression did not change in response to light stress. Bonisteel concluded with the hypothesis that LL Prochlorococcus may be allocating their energy resources to other processes that would advantage life under low light conditions.

Table 1. FtsH protease homologs amongst the model cyanobacterium Synechocystis sp. PCC6803 and Prochlorococcus marinus. Protein homology determined by multiple sequence alignment with MUSCLE followed by construction of maximum likelihood phylogenetic tree using 1000 bootstrap replicates in MEGAX

Organism	Protein Homolog			
Prochlorococcus marinus	FtsH1	FtsH2	FtsH3	FtsH4
Synechocystis sp. PCC6803	Slr0228	Slr1604	Sll1463	Slr1390
k
Function

This correlates with the above research on photosystem composition in laboratory cultivated HL and LL Prochlorococcus. HL strains are found primarily in oligotrophic surface waters where light is abundant, thus allocate more of their energy to maintaining their photosynthetic apparatuses and can obtain energy from the reduction of sugars. In contrast, LL strains live under eponymically low irradiances and may depend on environmental nutrients and cyclic electron flow to maintain energy homeostasis.

@@

```{r MED4PAR, fig.cap = MED4PAR_cap, fig.height = 6, fig.width = 8, echo = FALSE}

```{r SS120PAR, fig.cap = SS120PAR_cap, fig.height = 6, fig.width = 8, echo = FALSE}

```{r MIT9313PAR, fig.cap = MIT9313PAR_cap, fig.height = 6, fig.width = 8, echo = FALSE}

@@

@@

# Acknowledgements
Jonah Sheinin assisted with growth of some cultures for protein analyses.

# Funding
To Be Entered through PLoSONe system and deleted here
Czech Academy of Science visiting fellowship (DAC)
Canada Research Chair in Phytoplankton Ecophysiology (DAC)
Natural Sciences and Engineering Research Council of Canada, 'Latitude and Light' (DAC)
Canada Foundation for Innovation (DAC)
New Brunswick Foundation for Innovation (DAC)
Rice Graduate Fellowship (MS)
